Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-02-05 paper1_jiturra.Rmd
up

always_allow_html: true      
linkcolor: black                         # enlaces y citas en color azul
bibliography: input/bib/rgroup-redist.bib     # bibliografia en bibtex
csl: input/bib/apa6.csl
editor_options:
  chunk_output_type: console            # en RStudio, mostrar output en consola
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm" # márgenes de página
header-includes:
  - \usepackage{times}           # Times New Roman
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \setlength\parindent{24pt} 
---

@@

```{r setup, include=FALSE}
if (!require("pacman")) install.packages("pacman")  #si falta pacman, instalar
if (!require("tinytex")) install.packages("tinytex")#si falta tinytex, instalar
pacman::p_load(knitr, kableExtra, dplyr, ggplot2,sjmisc,texreg) # librerias
knitr::opts_chunk$set(warning = FALSE,  # mensaje de warning
                      message = FALSE,  # mensajes/avisos de librerias  
                      cache = F,    # cache de los chunks,usar analisis pesados
                      out.width = '100%',
                      fig.pos= "H",     # posicion figuras H = HERE
                      fig.align = "center",
                      echo = FALSE      # incluir chunk en output
                      )
options(scipen=999) # notacion cientifica
rm(list=ls())       # limpiar workspace
options(knitr.kable.NA = '') # NA en kable = ''
```

# Introduction

@@

Theoretically, self-interest has served as the canonical model to explain redistributive preferences, where the material interests of those with the most to gain from redistribution conflict with those in positions of greater material well-being [@meltzer_rational_1981]. Similarly, extensions of the model have paid more attention to occupation as a representation of potential unemployment risk, arguing that social classes with greater risk exposure are prone to support redistributive policies as they reduce labor market insecurity [@rehm_risks_2009]. Differently, part of the scholarly discussion has stressed the role of cognitive processes in preference formation for understanding why, even in contexts of greater economic disparities, the lower classes oppose redistributive policies, arguing that misinformation plays a crucial role in linking material conditions with the redistributive policies [@bartels_homer_2005;@druckman_preference_2000]. Similarly, it has been argued that contexts of greater segregation affect the inferences about social and economic inequality [@summers_deliberating_2022;@mijs_inequality_2018] that ultimately contribute to the decoupling of material interests from economic preferences [@hvidberg_social_2020;@garcia-castro_changing_2022].

As material interests might prevail in scarcity, normative or value-driven motivations emerge as an alternative explanation in preference formation when pure self-interest weakens. Thus, values such as self-transcendence or humanitarianism share a common spirit regarding their desire for equality and concern for human dignity, providing the common ground for altruism that ultimately motivates egalitarian preferences [@Kulin2013;@feldman_humanitarian_2001]. However, the protagonism of values is indeed conditioned to risk exposure and resource availability, becoming salient under conditions of greater security against market risks and material prosperity and weakening under conditions of greater vulnerability and material precariousness [@Maldonadoetal2019]. Related arguments have suggested that skills specificity and work logic represented in occupations play a role in forming political attitudes [@oesch_redrawing_2006]. Specifically, it is argued that the work logic, together with the performed tasks, socializes individuals given the time spent in their job as well as the provision of goods and services, suggesting that technical and managerial occupations hold more conservative political views compared to those performing interpersonal services [@oesch_electoral_2018]. However, it is challenging to disentangle if the class differences in attitudes are indeed explained through political socialization developed within workplaces because they might result from the internalization of social norms in early-life socialization that motivate self-selection into specific career paths or the organizational hiring logic requirements for interpersonal or managerial positions [@kitschelt_occupations_2014].

@@

As social relations within the occupational structure serve as the foundation for class analysis, widening the examination to other dimensions of social life, such as family dynamics, the development of friendships, and diverse social activities, can significantly enrich the scope of class analysis, particularly concerning its relation to attitude formation. In this sense, social stratification scholars have paid attention to the intersection of status positions and group membership as the observable patterns of association referred to as relational networks [@blau_inequality_1977]. Empirically, homophily is one of the most significant findings in network research, whereas the likelihood of tie formation between individuals of similar characteristics is strong [@mcpherson_birds_2001; @lazarsfeld_friendship_1954], such as age and gender in tracing friendship and family ties, as well as the segregated social environments alongside ethnicity and socioeconomic status [@bargsted_pautas_2020; @plaza_juntos_2022]. In addition, psychosocial factors such as socialization preferences and group dynamics also play a role in segregation processes, considering that isolated social environments decrease the interchange of world views, reinforcing preferences towards in-group members and polarizing attitudes [@visser_attitudes_2004; @diprete_segregation_2011]. Nonetheless, without denying the relevance of socialization preferences, segregation has been predominantly explained through the lens of social differentiation in social activities that work as foci of contact opportunities that ultimately consolidate social networks [@feld_focused_1981; @mcpherson_network_2019]. From this scope, we argue that class positions consider differences in economic or cultural resources and patterns of sociability observed in the distribution of social ties.

How classes are formed also provides insights regarding their social relations and sociability practices. In sociology, classes are understood as the everyday basis for action, given their labor market situation, that ultimately shapes their access to opportunities and resources [@weber_economy_1978]. According to habitus theory, classes incur a series of distinctive practices pivoted in cultural capital as the basis for mutual recognition [@bourdieu_distinction_1984]. Thus, classes seek to improve their position through the intergenerational transmission of resources, whereas similarities in friendship and family choices play a crucial role in the reproduction privilege. Additionally, while symbolic resources reinforce status distinctions, access to exclusive social activities in the upper class increases segregation at the expense of excluding the lower classes [@bourdieu_reproduccion_1981]. Ultimately, classes consolidate their demographic and cultural identities by mobilizing resources and shared sociability practices, ensuring intergenerational reproduction of their structural positions [@goldthorpe_sobre_1992]. Altogether, these processes further reinforce social distinctions between classes and contribute to the persistence of class structures. Hence, the socialization of common lifestyles and shared worldviews facilitates the consolidation of social classes. Access to social resources embedded in social networks follows a pattern of accumulation that reproduces pre-existing inequalities, in which class positions are vital in accounting for mechanisms of exclusion and segregation across the class structure.

Previous research on social class and tie formation has suggested that class permeability, understood as the formation of cross-class social ties, is traced differently across property and authority boundaries. Empirically, evidence from Western industrialized countries has shown that the formation of friendship ties between owners and the working class is very less likely, in contrast to the tendency of higher tie formation between supervisors and workers, suggesting that the authority dimension is much less permeable than authority-based boundaries given that class interests increase the social distance between proprietors and workers, while the intermediate class position of supervisors and contact frequency with workers make friendship tie formation more likely [@wright_relative_1992]. Similarly, evidence from Chile has shown that the higher permeability of the intermediate classes contrasts with the more homogeneous networks of the working class, suggesting that their limited resources and lower capacity to be socially engaged ultimately result in a lack of social resources that isolate them from others, while the upper class is less permeable and homogenous as a result of its tendency to self-selection as a practice that ultimately seeks to reproduce their privileged status positions [@otero_open_2021].

In contrast, cross-class embeddedness can be described regarding their social activities as well as through their changes across generations and time. In this regard, @pichler_social_2009 suggest that higher civic engagement in formal organizations increases the chances of bridging with diverse people among the upper class, in contrast to the working class, which tends to be more homogeneous in its civic engagement behavior. Similarly, other studies in Argentina and Croatia have exhibited that the upper and intermediate classes hold increasingly diverse and prestigious social environments than the working classes, where being socially mobile does help to improve both dimensions in tie formation, whereas it does not equalize the weight of class background compared to those intergenerationally stable in the upper class [@cepic_how_2020;@carrascosa_class_2023]. Additionally, the few longitudinal evidence suggests that networks change following a dynamic of cumulative advantages regarding its composition in terms of prestige and diversity, where upper classes improve in both dimensions while the lower class shows more stability across the life course [@volker_social_2020].

Theoretically, these studies account for class-based network segregation as the degree of connectedness of an individual to different occupations is meant to vertically represent social resources embedded in social networks [@lin_social_2007]. Drawing on this approach, the attempts of stratification scholars have been focused on how social connections are distributed across the class structure as a matter of social integration [@blau_macrosociological_1977]. Therefore, _homogeneity_  is described as the lack of cross-class network ties. It is indeed conceptually more proximate to the homophily principle because it is anchored in ego’s class position. At the same time, _diversity_ is defined as the rate of dissimilar ties or simply the total ties to certain groups that do not necessarily count with a reference position to describe the network composition. Hence, both approaches provide a set of alternatives for the study of social class from a network perspective that has been increasingly discussed in the stratification literature as well as in the scholarly discussion about social class and political attitudes.

>>>>

always_allow_html: true      
linkcolor: black                         # enlaces y citas en color azul
bibliography: input/bib/rgroup-redist.bib     # bibliografia en bibtex
csl: input/bib/apa6.csl
editor_options:
  chunk_output_type: console            # en RStudio, mostrar output en consola
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm" # márgenes de página
header-includes:
  - \usepackage{times}           # Times New Roman
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \setlength\parindent{24pt} 
  - \usepackage{fancyhdr}
  - \fancyhead{} 
---

@@

```{r setup, include=FALSE}
if (!require("pacman")) install.packages("pacman")  #si falta pacman, instalar
if (!require("tinytex")) install.packages("tinytex")#si falta tinytex, instalar
pacman::p_load(knitr, kableExtra, dplyr, ggplot2,sjmisc,texreg) # librerias
knitr::opts_chunk$set(warning = FALSE,  # mensaje de warning
                      message = FALSE,  # mensajes/avisos de librerias  
                      cache = F,    # cache de los chunks,usar analisis pesados
                      out.width = '100%',
                      fig.pos= "H",     # posicion figuras H = HERE
                      fig.align = "center",
                      echo = FALSE      # incluir chunk en output
                      )
options(scipen=999) # notacion cientifica
rm(list=ls())       # limpiar workspace
options(knitr.kable.NA = '') # NA en kable = ''
```
\pagestyle{fancy}
\fancyhead[R]{ \thepage}
\fancyhead[L]{Julio Iturra-Sanhueza}
# Introduction

@@

Theoretically, self-interest has served as the canonical model to explain redistributive preferences, where the material interests of those with the most to gain from redistribution conflict with those in positions of greater material well-being [@meltzer_rational_1981]. Similarly, extensions of the model have paid more attention to occupation as a representation of potential unemployment risk, arguing that social classes with greater risk exposure are prone to support redistributive policies as they reduce labor market insecurity [@rehm_risks_2009]. Differently, part of the scholarly discussion has stressed the role of cognitive processes in preference formation for understanding why, even in contexts of greater economic disparities, the lower classes oppose redistributive policies, arguing that misinformation plays a crucial role in linking material conditions with the redistributive policies [@bartels_homer_2005;@druckman_preference_2000]. Similarly, it has been argued that contexts of greater segregation affect the inferences about social and economic inequality [@mijs_inequality_2018] that ultimately contribute to the decoupling of material interests from economic preferences [@hvidberg_social_2020;@garcia-castro_changing_2022].

As material interests might prevail in scarcity, normative or value-driven motivations emerge as an alternative explanation in preference formation when pure self-interest weakens. Thus, values such as self-transcendence or humanitarianism share a common spirit regarding their desire for equality and concern for human dignity, providing the common ground for altruism that ultimately motivates political preferences [@Kulin2013;@feldman_humanitarian_2001]. Despite the protagonism of values, these are indeed conditioned to risk exposure and resource availability, becoming salient under conditions of greater security against market risks and material prosperity and weakening under conditions of greater vulnerability and material precariousness [@Maldonadoetal2019]. Besides, it has been argued that occupations play a role in political attitude formation as they represent a set of specific skills and work logic that convey particular normative views [@oesch_redrawing_2006]. Thus, workspaces are conceived as socializing agents, considering that economically active people spend a substantial part of their time at work, including the social relations involved in providing goods and services. Hence, empirical evidence has suggested that technical and managerial professionals hold more conservative political views than occupations in interpersonal services [@oesch_electoral_2018;@lindh_class_2020]. Despite this, it is challenging to disentangle if the class differences are indeed explained through work experiences because it is feasible to argue that political preferences are the result of early-life internalization of social norms, which might be crystalized in career self-selection paths, which also might be influenced by the organizational hiring requirements for specific jobs [@kitschelt_occupations_2014].

@@

As social relations within the occupational structure serve as the foundation for class analysis, widening the examination to other dimensions of social life, such as family dynamics, the development of friendships, and diverse social activities, can significantly enrich the scope of class analysis, particularly concerning its relation to attitude formation. In this sense, social stratification scholars have paid attention to the intersection of status positions and group membership as the observable patterns of association referred to as relational networks [@blau_inequality_1977]. Empirically, homophily is one of the most significant findings in network research, whereas the likelihood of tie formation between individuals of similar characteristics is strong [@mcpherson_birds_2001; @lazarsfeld_friendship_1954], such as age and gender in tracing friendship and family ties, as well as the segregated social environments alongside ethnicity and socioeconomic status [@bargsted_pautas_2020; @plaza_juntos_2022]. In addition, psychosocial factors such as socialization preferences and group dynamics also play a role in segregation processes, considering that isolated social environments decrease the interchange of world views, reinforcing preferences towards in-group members and polarizing attitudes [@visser_attitudes_2004; @diprete_segregation_2011]. Nonetheless, without denying the relevance of socialization preferences, segregation has been predominantly explained through the lens of social differentiation in social activities as foci of contact opportunities that ultimately consolidate social networks [@feld_focused_1981; @mcpherson_network_2019]. From this scope, we argue that class relations not only represent resources-based distinctions, but patterns of sociability and exchange observed in the differentiation of social ties.

How classes are formed also provides insights regarding their social relations and sociability practices. In sociology, classes are understood as the everyday basis for action, given their labor market situation, that ultimately shapes their access to opportunities and resources [@weber_economy_1978]. According to habitus theory, classes incur a series of distinctive practices pivoted in cultural capital as the basis for mutual recognition [@bourdieu_distinction_1984]. Thus, classes seek to improve their position through the intergenerational transmission of resources, whereas similarities in friendship and family choices play a crucial role in the reproduction privilege. Additionally, while symbolic resources reinforce status distinctions, access to exclusive social activities in the upper class increases segregation at the expense of excluding the lower classes [@bourdieu_reproduccion_1981]. Ultimately, classes consolidate their demographic and cultural identities by mobilizing resources and shared sociability practices, ensuring intergenerational reproduction of their structural positions [@goldthorpe_sobre_1992]. Hence, the socialization of common lifestyles and shared worldviews facilitates the consolidation of social classes, whereas social resources embedded in social networks follow a pattern of accumulation and exclusion that ultimately leads to social segregation in both the upper and the lower bounds of the social structure.

Previous research on social class and tie formation has suggested that class permeability, understood as the formation of cross-class social ties, is traced differently across property and authority boundaries. Empirically, evidence from Western industrialized societies has shown that the formation of friendship ties between owners and the working class is very less likely, in contrast to the tendency of higher tie formation between supervisors and workers, suggesting that the authority dimension is much less permeable than authority-based boundaries given that class interests increase the social distance between proprietors and workers, while the intermediate class position of supervisors and contact frequency with workers make friendship tie formation more likely [@wright_relative_1992]. Similarly, evidence from Chile has shown that the higher permeability of the intermediate classes contrasts with the more homogeneous networks of the working class, suggesting that their limited resources and lower capacity to be socially engaged ultimately result in a lack of social resources that isolate them from others, while the upper class is less permeable and homogenous as a result of its tendency to self-selection as a practice that ultimately seeks to reproduce their privileged status positions [@otero_open_2021].

In contrast, cross-class embeddedness can be described regarding their social activities as well as through their changes across generations and time. In this regard, @pichler_social_2009 suggest that higher civic engagement in formal organizations increases the chances of bridging with diverse people among the upper class, in contrast to the working class, which tends to be more homogeneous in its civic engagement behavior. Similarly, other studies in Europe and South America have exhibited that the upper and intermediate classes hold increasingly diverse and prestigious social environments than the working classes, where being socially mobile does help to improve both dimensions in tie formation, whereas it does not equalize the weight of class background compared to those intergenerationally stable in the upper class [@cepic_how_2020;@carrascosa_class_2023]. Additionally, the few longitudinal evidence suggests that networks change following a dynamic of cumulative advantages regarding its composition in terms of prestige and diversity, where upper classes improve in both dimensions while the lower class shows more stability across the life course [@volker_social_2020].

Theoretically, these studies account for class-based network segregation as the degree of connectedness of an individual to different occupations is meant to represent social resources embedded in social networks vertically in the social structure [@lin_social_2007]. Drawing on this approach, the attempts of stratification scholars have been focused on how social connections are distributed across the class structure as a matter of social integration [@blau_macrosociological_1977]. Besides, _homogeneity_  is described as the lack of cross-class network ties. It is indeed conceptually more proximate to the homophily principle because it is anchored in ego’s class position. At the same time, _diversity_ is defined as the rate of dissimilar ties or simply the total ties to certain groups that do not necessarily count with a reference position to describe the network composition. Hence, both approaches provide a set of alternatives for the study of social class from a network perspective that has been increasingly discussed in the stratification literature as well as in the scholarly discussion about social class and political attitudes.

## Network segregation and attitudes toward redistribution

Apart from the individual approach to attitude formation, we argue that people also form their opinions in the economic domain on the basis of their social relations. Despite the limited research on the link between social ties and attitudes in the economic domain, we identify two broad theoretical approaches that have discussed the role of social networks in attitude formation.

On the one hand, the _Reference Groups_ approach accentuates the role of beliefs about economic inequality rooted in social comparison processes as an explanation through which people form their opinions about redistributive policies [@condon_economic_2020]. This hypothesis can be traced to the studies on images of class and perceived class conflicts [@Evans1992;@KelleyEvans1995]. Here, the initial argument is that people form their beliefs through the individual, family, friends, and coworkers experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of reference groups [@Evans1992, p. 467]. From this perspective, how people infer the social world is linked to the degree of segregation in their immediate social environment, which influences the intensity and character of the information, ultimately shaping inequality perceptions [@mijs_is_2021]. Thus, conversations with friends, family, and work colleagues shape inequality beliefs through experience sharing. Interestingly, research employing collaborative games has shown that the socioeconomic diversity of participants in conversations contributes to reconstructing more accurate images of income and wealth inequalities than isolated individuals [@summers_deliberating_2022]. However, we argue that this research field has been more focused on the cognitive or informational dimension of preference formation through inequality perceptions [see @cansunar_who_2021;@garcia-castro_perceived_2022] or the role of informational treatments [see @Cruces2011;@mccall_exposure_2017;@becker_significant_2021], rather than empirically addressing the role of class segregation in social networks and its claimed influence on attitudes towards inequality.
